Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-11-20 Response_referees.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

List of major changes apported to the paper:
\begin{itemize}
 \item Much efforts have been spent in trying to clarify better the relation between localization properties and features of the computational basis needed 
 to provide precise results. This topic has been discussed in all the text. Major modifications under this perspective have been done in the introduction, in section II at pag. 5 before the analysis of fig. 2,in section IV and in the conclusions. 
 \item The introduction has been deeply revised. In particular we have reduced the length of the first two paragraph concerning the differences between accuracy and precision.
 \item We added a comment after eq. (7) on the non-hermiticity of the response density and corrected the equation (11).
 \item Added a dedicated subsection (actually the IIA at page 4)  to discuss the localization properties of the FS states for static perturbations. 
 \item Added a panel in figure 2 that contain the plot of the real part of $\alpha(\omega)$ in calculated in different computational domain. This new plot aims to evidence that the localization
 properties of th FS affect, on equal footing, both the real and the imaginary part of the linear response functional.
 \item We have reorganized the material after section III A. In particular we have introduced a new section (number IV in the actual version of the paper) in which we have discussed both the physical features of the excitation spectrum and the relative importance of localized vs. delocalized sector in expressing the susceptibility functional. This section also collects computational results for molecules beyond the case of $CO$ and benzene, already discussed in the previous version of the paper.  
\end{itemize}

@@

\section*{Response to the first referee}

We wish to thank the referee for the positive review on the paper. The text has been deeply revised according to the suggestions of both the referees.

Here we report some detailed answer concerning the main aspects raised in the referee's report:
\begin{enumerate}
  \item \emph{on the point concerning the dimension of the simulation box needed to achieve the convergence of the delocalized states of the $CO$ molecule}.

  Delocalized fluctuation states do not achieve convergence as long as the simulation box is increased. Since they behave as continuum states they suffer of the same continuum collapse behavior described in section IA for the unbound solution of the Schr\"odinger equation. What is expected to converge is the spectral function, or declined in the present case the real and imaginary part of the dynamical polarizability. 
  The value of convergence is system/observable and perturbation dependent. It can be found through an explicit convergence analysis by introducing an acceptance tolerance, however this procedure is out of the scope of our analysis and has not been performed.

  Nonetheless, to have an idea of the dimension of the box needed to achieve convergence we can refer to the case of the Benzene molecule and make usage
  of the result found in our analysis, compared to the one reporte in fig. 2 of \cite{baroni2008}. Within our analysis we show that a box linear size of about 18 Angostroem is sufficient to produce converged results for excitation spectrum below IP (around 6.5 eV in the present case). On the contrary, results of \cite{baroni2008} evidence that to achieve convergence up to 30 eV a box of linear dimension of more than 31 Angstroem is needed, so going from the localized to the delocalized sector of excitation spectrum requires a relevant increase in the simulation domain.

  \item \emph{on the point concerning the convergence rate of the delocalized vs. localized states with respect
  to the grid spacing}

 In presenting the results of this paper care has been taken to ensures the convergence of the results wrt the computational parameters. We have found that there are no relevant difference in the convergence rate of localized vs delocalized quantities (for instance excitations) in function of the grid spacing. This is due to the fact that localized and delocalized states differ in their asymptotic properties whereas the grid spacing convergence is influenced by the fast short-range oscillation of the functions.

 Add a comment in the text somewhere?

 The values of the grid spacing used for the calculations have been added in appendix C.

\item Other minor corrections:
\begin{itemize}
 \item The insets of fig. 4 and 5 have been increased.
 \item the \emph{reasonable behavior of the perturbed potential} at the end of section IIIA has been clarified. It simply means that the
 the perturbed potential $\op V'[\op E]$ gives rise to a bound state when applied to $\ket{\psi_p}$.
 \item explicit reference to the inset of fig. 5 has been performed in the text.
\end{itemize}

@@

\end{enumerate}

\section*{Response to the second referee}

We thank the referee for its comment on the paper. Thank to his/her advice and criticism we have tried to clarify our message. 
We try to answer to the critical points raised out by the review:

@@

 The first one concerns with the removal of the contribution of the delocalized states in a specific excitation. This

 The second one indicate what happens when one uses a basis able to express localized states only...

\end{enumerate}

\bibliographystyle{elsarticle-num}
\bibliography{Analytic_biblio}

### 2018-11-20 AnalyticStructure_main.tex
text revised and update in the referees report

%\usepackage[markup=underline]{changes}
%\usepackage[final]{changes}
%\definechangesauthor{dal}
%\definechangesauthor{gen}

@@

\subsection{Localization propeties at finite $\omega$. Emergence of a strong and a weak regime}

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$,
while they behave as unbound states for values of $\omega$ above these threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response
density in terms of FS implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;.
\ee
% \begin{multline}\lb{LinearResponseFunctDef2}
% \braket{\delta\op O}_\Phi(\omega) = \sum_p \left( \bra{f_p^\Phi(-\omega)}\op O\ket{\psi_p}+ \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \\
% =2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;.
% \end{multline}
In the second equality we have assumed that $\op O$ is a real symmetric operator and chosen a real set of occupied molecular states.
Let us suppose that also the observable of interest $\op O$ can be restricted to localized domains in the same way as the potentials $ \op V$ and $\op \Phi$.
Once again, it would be enough to consider a local operator for this condition to be met.

@@

It is possible to reformulate this type of analysis by directly looking at LR quantities that are defined independently from the perturbing operators and target observable. This can be achieved by studying the \emph{susceptibility functional}, defined implicitly trough the operator equation
\be\lb{ExcitationOperatorsDef1}
\Liouv \op E_a = \Omega_a \op E_a \qq \op{\tilde E}_a \Liouv = \Omega_a \op{\tilde E}_a \;,
\ee
together with the operator orthonormalization condition %$\trace{\op{\tilde E}\op E'} = \delta(E-E')$.
\be\lb{orthoExcitatioOpDef1}
\trace{\op{\tilde E}_a\op E_b} = \delta_{ab} \;.
\ee
Excitations defined in that way may be considered as a basis of the operators that satisfy the transverse condition as in
Eq.~\eqref{RhopTransverseDef1}, and enable us to express the susceptibility as a spectral decomposition:
\be\lb{RhopExcitationDef1}
\sop{{\cal \chi}}(\omega) \cdot   = \sum_{\{a\}} \op E_{a} \,
\frac{\trace{\commutator{\op{\tilde E}_a}{\dmnot}\cdot}}{\omega-\Omega_a} \;.
\ee
In this way Eq.~\eqref{LinearResponseFunctDef2} may be also written as:
\be\lb{deltaoexc}
  \braket{\delta \op O} =  \sum_{\{a\}} \frac{\trace{\op O \sop B_a \op \Phi(\omega)}}{\omega -\Omega_a}\;,
\ee
where the action of the spectral operator $\sop B_a$ reads
\be
  \trace{\op O \sop B_a \op \Phi(\omega)} = \trace{\op O \op E_a}
  \trace{\op{\tilde E}_a \commutator{\dmnot}{\op \Phi(\omega)}} \;. \nn
\ee

@@

The properties of the Liouvillian superoperator $\Liouv$ (sketched in appendix \ref{LiouvillianAction}) imply that both the left and right operator-valued
eigenstates \eqref{ExcitationOperatorsDef1} satisfy the transverse condition \eqref{RhopTransverseDef1} and can be parametrized as
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;.
\end{align}
Each excitation ``mode'' of the system, with associated energy $\Omega_a$, is thus described by a set of states $\{\phi^a_p,\chi^a_p\}_{p \in a}$, defined in the unoccupied subspace. These objects represent, respectively, the state in which $\ket{\psi_p}$ is excited -- or from whom it decays -- when the system is subject to the \emph{monochromatic} perturbation $\op \Phi_a \equiv \commutator{\dmnot}{\op E_a}$. We denote with the shorthand $p \in a$ the fact that the sum of Eq.~\eqref{ExcitationOperatorsDef2} is in general restricted only to a $a$-dependent subset of the occupied orbitals, depending on the particular symmetry
of the excitation. Excited states  satisfy the normalization condition
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;.
\ee
By applying \eqref{ExcitationOperatorsDef1} to the above parametrization
of the excitations, we obtain the following Sternheimer-like equations for the excited states:
\begin{align}\lb{ExcitationOperatorsDef3}
&\left[\Omega_a-(\hnot - \eps_p)\right] \ket{\phi_p^a} = \op Q_0 \op V'[\op E_a] \ket{\psi_p} \;, \nn\\
&\bra{\chi_p^a}\left[-\Omega_a-(\hnot - \eps_p)\right] = \bra{\psi_p} \op V'[\op E_a] \op Q_0  \;.
\end{align}
By writing these equations in a basis set of virtual states we obtain the well-known Casida's eigenvalue equation\cite{CasidaBook}
associated to the excitation energy $\Omega_a$ (see appendix \ref{casida}).
The spectrum is symmetric w.r.t the inversion of the eigenvalues $\Omega_a \rightarrow -\Omega_a$ and, given a specific excitation $\{\phi^a_p,\chi^a_p\}$, 
the associated solution with opposite energy is described by the transposed pair $\{\chi^a_p,\phi^a_p\}$.
We concentrate our analysis on the sector of positive energies, i.e. $\Omega_a > 0$.

@@

\section{Physical relevance of the excitation spectrum. A comparison among localized and continuum sectors}
% \deleted[id=dal]{
% Comments on the features and the physical relevance of the two sector of the excitation spectrum}
% 
% %\subsection{Localized sector}
% 
% %\begin{itemize}
% % \deleted[id=dal]{\item It is composed by a genuine set of discrete elements.}
% % \deleted[id=dal]{\item The values of the excitation energies do not depend on the computational setup (observable features) and so can be compared among different approaches.}
% % \deleted[id=dal]{\item The knowledge of this sector of the excitation spectrum allows us to express only the imaginary part of (proper equation that expresses the response functional written using Sokowsky-Plemely
% % on the susceptibility functional).}
% % \deleted[id=dal]{\item A computational setup in which delocalized degrees of freedom are excluded by design can in principle describe to this sector in a exact way.}
% %\end{itemize}
% 
% \subsection{Continuum sector}
% 
% \begin{itemize}
%    \item Make some statements on the features that a computational setup must have to guarantees that this sector provides converged results. Difference between systematic and localized basis setup. Can we comment about the needs to add ghost atoms for the evaluation of statical polarizability with gaussian basis?
%  % \deleted[id=dal]{\item Not true poles of the susceptibility (but only an effective tool to compute the spectral function).}
%  % \deleted[id=dal]{\item Deeply affected by the computational setup used to represent them. In particular, the presence of a finite computational domain implies a fictitious discretization of the excitation levels and the associated spectrum exhibits an explicit dependence from the size of the domain.}
%  % \deleted[id=dal]{\item This sector of the excitation spectrum is needed to compute imaginary parts above IP and real part for \emph{all} the values of $\omega$. Add examples on the statical polarizability. Can we comment}
% \end{itemize}
% 
% 
% \vspace{1cm}

@@

In Fig.~\ref{CO_exc} and \ref{C6H6_exc} we plot the DOS of the Liouvillian for the $CO$  and Benzene molecules,
for two different numerical setups of increasing size of the simulation domain.
The figures show clearly that all the excitations of energy $\Omega_a$ lying below the
molecule ionization potential (IP) have a localized behavior, as their energy does not vary with the size of the simulation domain.
On the other hand, above the first ionization threshold the excitations have an energy that strongly depends on the computational
setup, which is in line with the expected pseudo-continuum character of this sector.
In addition, for the benzene molecule it is also possible to
identify excitations of energy above IP that still possess a localized behavior, fulfilling the condition $\Omega_a < |\eps_a|$.
In Fig.~\ref{c6h6_excLand} we collect the excitations of the benzene molecule following this criterion:
the agreement with the expected analytic structure presented in Fig.~\ref{ExcitationLandscape} is remarkable.

\subsection{Relative importance of discrete and continuum excitations}
% \deleted[id=gen]{
% We have presented formal arguments, confirmed by numerical calculations, that show that
% the excitations belonging to the discrete sector of the spectrum have energies
% that can be \emph{quantitatively} compared between codes, thanks to their
% localized character.
% Such objects are LR quantities, associated to the molecule, whose numerical convergence
% can be found by employing techniques which can be borrowed from the ones adopted for GS calculations. Being expressed by localized excited states,
% computational setups with sufficient level of completeness are able to
% express with high \emph{precision} such section of the excitation spectrum.}

So far we have showed with both formal arguments and numerical calculations that the excitation spectrum split into two sectors with very different features.
It is so reasonable to ask which is the relative importance of these two sectors and in particular if there is the possibility to express observable quantities
in a suitable $\omega$ range, by only employing the sector of genuinely localized excitations.
% We may ask which is the relative importance of the two sectors and in particular if there is the possibility to express observable quantities
% in a suitable $\omega$ range, by only employing the sector of genuinely localized excitations.
Stated otherwise, will the \emph{restriction} of $\sop \chi(\omega)$ to only localized excitations can be useful to extract some physical quantities?

In this regard it is illustrative to apply the above arguments to the evaluation of the dynamical polarizability tensor $\alpha_{ij}(\omega)$.
Employing the formalism of Eq.~\eqref{deltaoexc} provides
\be%gin{align}
  \mathrm{Im}\left(\alpha_{ij}(\omega) \right) =
  \sum_{a} \trace{\mathbf r_i \sop B_a \mathbf r_j} \delta(\omega - \Omega_a)\lb{imalpha} \;.
  %\\
  % &= 2 \sum_p \bra{\psi_p} \mathbf r_i \left(\mathrm{Im}\ket{f_p^{\mathbf r_j}(\omega)}\right)\lb{imfs}\;,
\ee%nd{align}
The imaginary part of the dynamical polarizability tensor therefore can be seen as the partial density of states of the Liouvillian projected on
$\op{\mathbf r}_i\op{\mathbf r}_j$. For this reason, it is evident that below the \emph{first} excitation threshold (IP), the contributions to $\mathrm{Im}\left(\alpha_{ij}(\omega)\right)$ can be expressed only in terms of observable excited states. On the contrary, the non-observable, basis dependent continuum sector of the excitation starts to contribute to $\mathrm{Im}\left(\alpha_{ij}(\omega)\right)$ above IP; the localized sector is not anymore sufficient to express Eq.~\eqref{imalpha} in this $\omega$ range.

% Not surprisingly this behaviour is in coherency with the
% localization properties of the FS below IP; Eq.~\eqref{imfs} show that
% for its imaginary part.

%
% The imaginary part of the FS may be decomposed in two quantities:
%
%
% \deleted[id=gen]{
% Following the considerations of Sec.~\ref{FluctuationState}, we know that
% for $\omega$ below the IP threshold all these quantities may be expressed
% via asymptotically localized states.
% We have also seen in Sec.~\ref{FluctuationState} that
% there exist a sector where the fluctuation states are genuinely localized, thereby presenting reproducibility features. Therefore,
% knowing that $\ket{f_p^\Phi(\omega)} =\sop \chi(\omega) \op \Phi(\omega) \ket{\psi_p}$,
% it would be interesting to obtain information on whether the discrete, localized sector of the excitations which appear in the spectral decomposition of $\sop \chi(\omega)$
% will be sufficient to express a localized FS.
% Considering Eq.~\eqref{SusceptibilityDef1} and
% expressing the spectral decomposition of the linear susceptibility we find:}
% \be
% \mathrm{Im}\left(\ket{f_p^{\mathbf r_j}(\omega)} \right)= \fscd{D} + \fscd{C}
% \lb{FSExcitationDef1} \;,
% \ee
% with
% \begin{align}
%  \fscd{D} &= \sum_a^{|\Omega_a| < |\eps_a|} \ket{w_p^a} \bra{\psi_p} \op {\mathbf r}_j \ket{w_p^a} \delta\left(\omega-\Omega_a\right) \;, \nn \\
%  \fscd{C}
%  &= \int_{|\Omega_a| > |\eps_a|} \!\!\!\!\!\!\!\!\!\!\!\!\!\dd a \ket{w_p^a}  \ket{w_p^a} \bra{\psi_p} \op {\mathbf r}_j \ket{w_p^a} \delta\left(\omega-\Omega_a\right) \;, \nn
% \end{align}
% where %$\varphi_p^a[\op \Phi(\omega)] \equiv \bra{\psi_p}\op\Phi(\omega)\ket{w_p^a}$, and
% $\ket{w_p^a}=\ket{\phi_p^a}+\ket{\chi_p^a}$.
%
% In the above equation we have explicitly separated the contributions of the discrete ($D$) and continuous ($C$) excitations
% in the spectral decomposition \eqref{SusceptibilityDef1}.
% \deleted[id=gen]{The summation and the integration run only on the positive values of $\Omega_a$.}
% The states $\ket{w_p^a}$ that appear in the discrete sum
% or in the integral are genuinely localized and delocalized, respectively.
% For \emph{any} value of $\omega$ the ket $\fscd{D}$ has, by definition, a bound-state like
% character as it is expressed as a sum of localized contributions.
%
% It is easy to see that for $\omega$ below IP  we have $\fscd{C}=0$.
% Below threshold, continuum excitations \emph{do not} contribute to the imaginary part of the
% dynamical polarizability.

Let us now consider the features of the response density operator of Eq.~\eqref{rhoPrimeFluctuationStateDef1}.
The fluctuation state in this case can be written as
\be\lb{fsinexc}
\ket{f_p^{\mathbf r_j}(\omega)}=
\sop \chi(\omega) \op{\mathbf r_j} \ket{\psi_p}
= \sum_a \frac{\bra{w_p^a} \op{\mathbf r_j} \ket{\psi_p}}{\omega^2 - \Omega_a^2}
\ket{w_p^a}\;.
\ee
Here we have employed the notation $\ket{w_p^a}=\ket{\phi_p^a}+\ket{\chi_p^a}$.
The above equation shows that for \emph{any} value of $\omega$ both sectors of localized and delocalized states $\ket{w_p^a}$ contribute.
However the discussion of Sec.~\ref{FluctuationState} has shown that
the FS is a genuinely localized state below the IP threshold.
The equivalence of the Sternheimer formalism with the spectral representation of the LR susceptibility implies that in this regime the contribution of the delocalized excitations in Eq.~\eqref{fsinexc} has to resum into a localized, asymptotically vanishing FS.
Such a locality is evident for the imaginary part of the FS, which
can be straightforwardly restricted to the sum over $\ket{w_p^a}$ below IP (see Eq.~\eqref{imalpha}). For the real part of FS the \emph{entire} spectrum
of excitations contribute.

Hence, contrary to GS, \emph{for LR calculation,
even when the response density is localized, the contribution coming from the
continuum sector of the excitations can never be neglected}.
In other terms, we \emph{cannot} exclude that a fluctuation state, even if genuinely localized, will have a nonzero projection
in the essential part of the Liouvillian spectrum.
This fact shows that it is \emph{impossible} to restrict the spectral representation
of $\chi(\omega)$ to only localized, discrete, \emph{code-independent} excitations, even for representing LR results in the $\omega=0$ limit.

>>>>

@@

\subsection{Localization properties at finite $\omega$. Emergence of a strong and a weak regime}

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$,
while they behave as unbound states for values of $\omega$ above these threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response
density in terms of FS implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;,
\ee
% \begin{multline}\lb{LinearResponseFunctDef2}
% \braket{\delta\op O}_\Phi(\omega) = \sum_p \left( \bra{f_p^\Phi(-\omega)}\op O\ket{\psi_p}+ \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \\
% =2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;.
% \end{multline}
here we have assumed that $\op O$ is a real symmetric operator and chosen a real set of occupied molecular states.
Let us suppose that also the observable of interest $\op O$ can be restricted to localized domains in the same way as the potentials $ \op V$ and $\op \Phi$.
Once again, it would be enough to consider a local operator for this condition to be met.

@@

It is possible to reformulate this type of analysis by directly looking at LR quantities that are defined independently from the perturbing operators and target observable. This can be achieved by studying the \emph{susceptibility functional}, defined implicitly trough the operator equation
\be\lb{SusceptibilityDef1}
\dm'(\omega) = \left(\omega -\mathcal L \right)^{-1} \commutator{\op \Phi(\omega)}{\dmnot} = \sop{{\cal \chi}}(\omega) \op\Phi(\omega) \;.
\ee
The susceptibility (super)operator $\cal \chi(\omega)$ is therefore closely related to the resolvent of the Liouvillian of Eq.~\eqref{LiouvillianRhopomegaDef1}, which is completely defined
from the molecule Hamiltonian $\hnot[\dm]$.
The ``excitation modes''  of the molecule are defined through the \emph{excitation operators}, satisfying
\be\lb{ExcitationOperatorsDef1}
\Liouv \op E_a = \Omega_a \op E_a \qq \op{\tilde E}_a \Liouv = \Omega_a \op{\tilde E}_a \;,
\ee
together with the operator orthonormalization condition %$\trace{\op{\tilde E}\op E'} = \delta(E-E')$.
\be\lb{orthoExcitatioOpDef1}
\trace{\op{\tilde E}_a\op E_b} = \delta_{ab} \;.
\ee
Excitations defined in that way may be considered as a basis of the operators that satisfy the transverse condition as in
Eq.~\eqref{RhopTransverseDef1}, and enable us to express the susceptibility as a spectral decomposition
\be\lb{RhopExcitationDef1}
\sop{{\cal \chi}}(\omega) \cdot   = \sum_{\{a\}} \op E_{a} \,
\frac{\trace{\commutator{\dmnot}{\op{\tilde E}_a}\cdot}}{\omega-\Omega_a} \;.
\ee
In this way Eq.~\eqref{LinearResponseFunctDef2} may be also written as:
\be\lb{deltaoexc}
  \braket{\delta \op O} =  \sum_{\{a\}} \frac{\trace{\op O \sop B_a \op \Phi(\omega)}}{\omega -\Omega_a}\;,
\ee
where the action of the spectral (super)operator $\sop B_a$ reads
\be
  \sop B_a \cdot = \op E_a \trace{\commutator{\dmnot}{\op{\tilde E}_a} \cdot} \;. \nn
\ee

% \be
%   \trace{\op O \sop B_a \op \Phi(\omega)} = \trace{\op O \op E_a}
%   \trace{\op{\tilde E}_a \commutator{\dmnot}{\op \Phi(\omega)}} \;. \nn
% \ee

@@

The properties of the Liouvillian superoperator $\Liouv$ (sketched in appendix \ref{LiouvillianAction}) imply that both the left and right operator-valued
eigenstates \eqref{ExcitationOperatorsDef1} satisfy the transverse condition \eqref{RhopTransverseDef1} and can be parametrized as
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;.
\end{align}
Each excitation ``mode'' of the system, with associated energy $\Omega_a$, is thus described by a set of states $\{\phi^a_p,\chi^a_p\}_{p \in a}$, defined in the unoccupied subspace. These objects represent, respectively, the state in which $\ket{\psi_p}$ is excited -- or from whom it decays -- when the system is subject to the \emph{monochromatic} perturbation $\op \Phi_a \equiv \commutator{\op E_a}{\dmnot}$. We denote with the shorthand $p \in a$ the fact that the sum of Eq.~\eqref{ExcitationOperatorsDef2} is in general restricted only to a $a$-dependent subset of the occupied orbitals, depending on the particular symmetry
of the excitation. Excited states  satisfy the normalization condition
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;.
\ee
By applying \eqref{ExcitationOperatorsDef1} to the above parametrization
of the excitations, we obtain the following Sternheimer-like equations for the excited states:
\begin{align}\lb{ExcitationOperatorsDef3}
&\left[\Omega_a-(\hnot - \eps_p)\right] \ket{\phi_p^a} = \op Q_0 \op V'[\op E_a] \ket{\psi_p} \;, \nn\\
&\bra{\chi_p^a}\left[-\Omega_a-(\hnot - \eps_p)\right] = \bra{\psi_p} \op V'[\op E_a] \op Q_0  \;.
\end{align}
By writing these equations in a basis set of virtual states we obtain the well-known Casida's eigenvalue equation\cite{CasidaBook}
associated to the excitation energy $\Omega_a$ (see appendix \ref{casida}).
The spectrum is symmetric w.r.t the inversion of the eigenvalues $\Omega_a \rightarrow -\Omega_a$ and, given a specific excitation $\{\phi^a_p,\chi^a_p\}$, 
the associated solution with opposite energy is described by the transposed pair $\{\chi^a_p,\phi^a_p\}$.
We concentrate our analysis on the sector of positive energies, i.e. $\Omega_a > 0$.

@@

\section{Physical relevance of the excitation spectrum. A comparison among localized and continuum sectors}

@@

In Fig.~\ref{CO_exc} and \ref{C6H6_exc} we plot the DOS of the Liouvillian for the $CO$  and Benzene molecules,
for two different numerical setups of increasing size of the simulation domain.
The figures show clearly that all the excitations of energy $\Omega_a$ lying below the
molecule ionization potential (IP) have a localized behavior, as their energy does not vary with the size of the simulation domain.
On the other hand, above the first ionization threshold the excitations have an energy that strongly depends on the computational
setup, which is in line with the expected pseudo-continuum character of this sector.
In addition, for the benzene molecule it is also possible to
identify excitations of energy above IP that still possess a localized behavior, fulfilling the condition $\Omega_a < |\eps_a|$ (see the inset of 
Fig.~\ref{C6H6_exc}). In Fig.~\ref{c6h6_excLand} we collect the excitations of the benzene molecule following this criterion:
the agreement with the expected analytic structure presented in Fig.~\ref{ExcitationLandscape} is remarkable.

\subsection{Relative importance of discrete and continuum excitations}

So far we have showed with both formal arguments and numerical calculations that the excitation spectrum split into two sectors with very distinct features.
It is then reasonable to ask which is their relative importance and, in particular, if there is the possibility to express observable quantities, in a suitable $\omega$ range, by only employing the sector of genuinely localized excitations.
Stated otherwise, will the \emph{restriction} of $\sop \chi(\omega)$ to only localized excitations can be useful to extract some physical quantities?

In this regard it is illustrative to apply the above arguments to the evaluation of the dynamical polarizability tensor $\alpha_{ij}(\omega)$.
Employing the formalism of Eq.~\eqref{deltaoexc} provides
\be\lb{imalpha}
\mathrm{Im}\left(\alpha_{ij}(\omega) \right) =
\sum_{a} \trace{\mathbf r_i \sop B_a \mathbf r_j} \delta(\omega - \Omega_a) \;.
\ee
The imaginary part of the dynamical polarizability tensor therefore can be seen as the partial density of states of the Liouvillian projected on
$\op{\mathbf r}_i\op{\mathbf r}_j$. For this reason, it is evident that below the \emph{first} excitation threshold (IP), the contributions to $\mathrm{Im}\left(\alpha_{ij}(\omega)\right)$ can be expressed only in terms of observable excited states. On the contrary, the non-observable, basis dependent continuum sector of the excitation starts to contribute to $\mathrm{Im}\left(\alpha_{ij}(\omega)\right)$ above IP; the localized sector is not anymore sufficient to express Eq.~\eqref{imalpha} in this $\omega$ range.

To further address this point we exploit the formal relation between response density and susceptibility functional %established in ~\eqref{SusceptibilityDef1} 
$\dm'(\omega) = \sop \chi(\omega)\op\Phi(\omega)$ and make usage of the diadic representation ~\eqref{rhoPrimeFluctuationStateDef1} of the response density. Pursuing this procedure provides the excitation-based
representation of FS as follows:
% Let us now consider the features of the response density operator of Eq.~\eqref{rhoPrimeFluctuationStateDef1}.
% The fluctuation state in this case can be written as
% \be\lb{fsinexc}
% \ket{f_p^{\mathbf r_j}(\omega)}=
% \sop \chi(\omega) \op{\mathbf r_j} \ket{\psi_p}
% = \sum_a \frac{\bra{w_p^a} \op{\mathbf r_j} \ket{\psi_p}}{\omega^2 - \Omega_a^2}
% \ket{w_p^a}\;,
% \ee
\be\lb{fsinexc}
\ket{f_p^{\mathbf r_j}(\omega)}=
\sum_a \frac{\bra{w_p^a} \op{\mathbf r_j} \ket{\psi_p}}{\omega^2 - \Omega_a^2}
\ket{w_p^a}\;,
\ee
where we have employed the notation $\ket{w_p^a}=\ket{\phi_p^a}+\ket{\chi_p^a}$.
%The above equation shows that for \emph{any} value of $\omega$ both sectors of localized and delocalized states $\ket{w_p^a}$ contribute.
The above equation shows that both the sectors of localized and delocalized states $\ket{w_p^a}$ contribute to the FS for \emph{any} value of $\omega$.
However the discussion of Sec.~\ref{FluctuationState} has shown that the FS is a genuinely localized state below the IP threshold.
The equivalence of the Sternheimer formalism with the spectral representation of the LR susceptibility implies that in this regime the contribution of the delocalized excitations in Eq.~\eqref{fsinexc} has to resum into a localized, asymptotically vanishing FS.
% Such a locality is evident for the imaginary part of the FS, which can be straightforwardly restricted to the sum over $\ket{w_p^a}$ below IP (see Eq.~\eqref{imalpha}). For the real part of FS the \emph{entire} spectrum
% of excitations contribute.
The localization below threshold is manifest for the imaginary part of the FS, that is explicitly written as a the sum over $\ket{w_p^a}$ below IP (see Eq.~\eqref{imalpha}). Instead, for the real part of FS, the \emph{entire} excitation spectrum contributes to the construction of the FS and the localization
\emph{has} to realize due to a non trivial cancellation of the unbound parts of the $\ket{w_p^a}$'s.

Hence, contrary to GS, we conclude that \emph{for LR calculation, even when the response density is localized, the contribution coming from the
continuum sector of the excitations can never be neglected}. In other terms, we \emph{cannot} exclude that a fluctuation state, even if genuinely localized, will have a nonzero projection in the essential part of the Liouvillian spectrum. This fact shows that it is \emph{impossible} to restrict the spectral representation of $\chi(\omega)$ to only localized, discrete, \emph{code-independent} excitations, even for representing LR results in the $\omega=0$ limit.

\subsection{Effective representation of $\sop \chi(\omega)$ for localized response densities}
The above arguments have shown that the only quantity which can be expressed in terms of the localized sector of the excitation spectrum is the imaginary part of the susceptibility functional $\sop \chi(\omega)$ in the optical regime (i.e. $\omega <$  IP). The response density for a generic perturbation and $\omega$ value requires the inclusion in the Liouvillian DOS of the delocalized part of the excitation spectrum, even in view of expressing localized fluctuation states.

A given numerical treatment typically provides only an effective description of the sector of delocalized excitation which in turns give rise to an effective 
representation of the susceptibility functional. It appears therefore important to identify indicators that enable us to validate the capability
of $\sop \chi(\omega)$ to express the localised FS (and thus the complete response density $\dm'_\Phi(\omega)$) below the IP threshold.
%It appears therefore important to identify indicators that might assess the capability of
%$\sop \chi(\omega)$ to express localised FS (and thus the complete response density $\dm'_\Phi(\omega)$) below the IP threshold.
For instance, the static polarizability tensor $\alpha_{ij}$ can be calculated as
% \be\lb{sminus2}
% \alpha_{ij} = \sint \dd a \frac{w^a_{ij}}{\Omega_a^2}\,, \quad w^a_{ij} \equiv \sum_{p\in a} \bra{\psi_p} \op {\mathbf r}_i \ket{w_p^a}\bra{\psi_p} \op {\mathbf r}_j \ket{w_p^a}\;.
% \ee
\be\lb{sminus2}
\alpha_{ij} = \sint \dd a \frac{w^a_{ij}}{\Omega_a^2}\,, \quad w^a_{ij} \equiv \sum_{p\in a} |\bra{\psi_p} \op {\mathbf r}_i \ket{w_p^a}|^2 \;.
\ee
Such value can be straightforwardly compared with results of Eq.~\eqref{staticalpha}, where the localized FS is obtained from GS techniques as
described in Sec.~\ref{FluctuationState}.
This is the well-know $S_{-2}$ sum-rule (see e.g.\cite{Wagner2012}), that express the value of the static polarizability
in terms of the molecule's oscillator strengths.
In the context of a numerical evaluation of the excitations, such sum rule can therefore interpreted as the capability of the pseudo-continuum sector of the excitation to
express the localized FS -- and consequently the reference value of $\alpha_{ij}$ -- at $\omega=0$.
